### 2014-08-21 src/docs/talks/msr-nyc-2014/stan-current-future.tex
added draft of talk for MSR NYC 2014 talk

\begin{document}
\sf%
\vspace*{6pt}
%
\noindent
\spc{\huge\bfseries \color{MidnightBlue}{Stan}}
\\[8pt]
\spc{\Large\bfseries \color{MidnightBlue}{Probabilistic Programming Language}}
\\
\vfill
\noindent
\spc\spc{\slshape\footnotesize Core Development Team:}
\\[2pt]
\spc\spc{\small Andrew Gelman, \  \myemph{Bob Carpenter}, \  Matt Hoffman}
\\
\spc\spc{\small Daniel Lee, \  Ben Goodrich, \  Michael Betancourt, \ }
\\
\spc\spc{\small Marcus Brubaker, \   Jiqiang Guo, \  Peter Li, \ }
\\
\spc\spc{\small Allen Riddell, \  Marco Inacio}
\vfill
\mbox{ } 
\hfill
\includegraphics[width=27.5pt]{img/stanlogo-main.pdf}\hspace*{9pt}
\\
\spc{\footnotesize Paris-Dauphine 2014}
\hfill 
{\footnotesize \tt mc-stan.org}

@@

\begin{itemize}
\item Stan is an \myemph{imperative} probabilistic programming language
\vspace*{-12pt}
\begin{itemize}\footnotesize
\item  cf., BUGS: declarative; \ Church: functional; \ Figaro: object-oriented
\end{itemize}
\item Stan \myemph{program}
\vspace*{-4pt}
\begin{itemize}\small
\item declares variables
\item codes log posterior (penalized likelihood)
\end{itemize}
\item Stan \myemph{inference}
\vspace*{-4pt}
\begin{itemize}\small
\item MCMC for full Bayesian inference
\item MLE for point estimation
\end{itemize}
\item Stan is \myemph{open source} \ {\footnotesize (BSD core C++, GPLv3 interfaces)}
\\
{\footnotesize hosted on GitHub; uses Eigen matrix lib, Boost C++ lib, googletest}
\end{itemize}

\sld{Who's Using Stan?}
\begin{itemize}
\item 750+ registrations for mailing list across the physical,
biomedical, and social sciences, plus engineering, finance and marketing

\item Subjects of Papers on Google Scholar in the sciences
\\
{\footnotesize clinical drug
trials, general computational statistics, entomology, opthalmology,
neurology, sociology and population dynamics, genomics, agriculture,
psycholinguistics, molecular biology, population dynamics, materials
engineering, botany, astrophysics, oceanography, election prediction,
fisheries, cancer biology, public health and epidemiology, population
ecology, collaborative filtering for recommender systems, climatology,
educational testing, natural language processing}
\end{itemize}

\sld{Books and Model Sets}
%
\begin{itemize}
\item 400+ page core language tutorial manual
\item separate installation, getting started, interface manuals 
\\ {\footnotesize  (RStan, PyStan, CmdStan)}
\item BUGS and JAGS examples (all 3 volumes), 
\item Gelman and Hill, {\slshape Data Analysis Using Regression and
    Multilevel/Hierarchical Models}
\item Wagenmakers and Lee, {\slshape Bayesian Cognitive Modeling}
\item two other books in progress
\end{itemize}

\sld{Scaling and Evaluation}
\begin{itemize}
\item Types of scaling
\vspace*{-4pt}
\begin{itemize}\small
\item more data
\item more parameters
\item \myemph{more complex models}
\end{itemize}
\item for MCMC, measure (as function of scaling)
\vspace*{-4pt}
\begin{itemize}\small
\item time to convergence
\item time per effective sample at mixing
\item memory usage
\end{itemize}
\end{itemize}

\sld{Stan vs. BUGS / JAGS}
\begin{itemize}
\item JAGS is the former \myemph{state-of-the-art}, \myemph{black-box} MCMC method
\\ {\footnotesize (Gibbs with slice sampling; chokes w.\ high posterior correlation)}
\item Stan (2.0) vs. BUGS on scaling IRT model
\vspace*{-4pt}
\begin{itemize}\footnotesize
\item Stan about \myemph{10 times faster}
\item Stan memory and time scales linearly
\item linearly scales with number of question answers
\end{itemize}
\vspace*{-3pt}
\vfill
\hspace*{-12pt}
\includegraphics[height=0.6in]{img/irt_scale-eps-converted-to.pdf}
\includegraphics[height=0.45in]{img/irt-memory-2.png}
\item Stan can handle problems that choke BUGS and JAGS
\end{itemize}

@@

@@

\begin{itemize}
\item \myemph{\tt\bfseries data} \ (once) 
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: declare data types, sizes, and constraints
\item {\slshape execute}: read from data source, validate constraints
\end{itemize}
%
\item \myemph{\tt\bfseries parameters} \ (every log prob eval)
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: declare parameter types, sizes, and constraints
\item {\slshape execute}: transform to constrained, Jacobian
\end{itemize}
%
\item \myemph{\tt\bfseries model} \ (every log prob eval) 
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: statements definining posterior density
\item {\slshape execute}: execute statements
\end{itemize}
\end{itemize}

@@

\sld{User-Defined Functions (Stan 2.3)}

@@

\begin{itemize}
\item Posterior \myemph{mode finding} via BFGS optimization
\\ {\footnotesize (uses model gradient, efficiently approximates Hessian)}
\item \myemph{Disables Jacobians} for parameter inverse transforms
\item Models, data, initialization as in MCMC
\vfill
\item \myemph{Very Near Future}
\vspace*{-4pt}
\begin{itemize}\small
\item  \myemph{Standard errors} on unconstrained scale
\\
{\footnotesize  (estimated using curvature of penalized log likelihood function}
\item Standard errors \myemph{on constrained scale})
\\
{\footnotesize  (sample unconstrained approximation and inverse transform)}
\item L-BFGS optimizer
\end{itemize}      
\end{itemize}

@@

\begin{itemize}
\item No-U-turn sampler (\myemph{NUTS})
\item Subtle algorithm to maintain \myemph{detailed balance}
\item Move randomly \myemph{forward or backward in time}
\item Double number of leapfrog steps each move (\myemph{binary tree})
\item Stop when a subtree makes a \myemph{U-turn}
\\ {\footnotesize (rare: throw away second half if not end to end U-turn)}
\item \myemph{Slice sample} points along last branch of tree
\item \myemph{Generalized} to Riemannian-manifold HMC
\\ {\footnotesize (Betancourt, arXiv)}
\end{itemize}

@@

\mypart{Part III}{What's Next?}

\sld{Differential Equation Solver}
\begin{itemize}
\item Auto-diff solutions w.r.t.\ parameters
\item Integrate coupled system for solution with partials
\item Auto-diff coupled Jacobian for stiff systems
\vfill
\item C++ prototype integrated for large PK/PD models
\vspace*{-4pt}
\begin{itemize}\footnotesize
\item 
Project with Novartis: 
  longitudinal clinical trial w.\ multiple drugs, dosings, placebo control,
  hierarchical model of patient-level effects, meta-analysis
\item
Collaborators: Frederic Bois, Amy Racine, Sebastian Weber
\end{itemize}
\end{itemize}

\sld{L-BFGS Optimizer}
\begin{itemize}
\item Faster, lower memory, more stable than BFGS
\item Plug-and-play in interfaces
\vfill
\item Code and doc complete; under testing
\end{itemize}

@@

\sld{Riemannian Manifold HMC}
\begin{itemize}
\item \myemph{NUTS} generalized to RHMC
\\ {\footnotesize (Betancourt {\slshape arXiv} paper)}
\item \myemph{SoftAbs} metric
\vspace*{-4pt}
\begin{itemize}\footnotesize
\item Eigendecompose Hessian
\item \myemph{positive definite} with positive eigenvalues
\item \myemph{condition} by narrowing eigenvalue range
\item Betancourt {\slshape arXiv}\ paper
\end{itemize}
\vfill
\item Code complete; awaiting higher-order auto-diff
\end{itemize}

\sld{Thermodynamic Sampler}
\begin{itemize}
\item Physically motivated alternative to ``simulated''
  \myemph{annealing and tempering} (not really simulated!)
\item Supplies external \myemph{heat bath}
\item Operates through \myemph{contact manifold}
\item System relaxes more naturally between energy levels
\vfill
\item Prototype complete
\\ {\small (Betancourt paper on {\slshape arXiv}; for geometers)}
\end{itemize}

\sld{Marginal Maximum Likelihood}
\begin{itemize}
\item Marginalize out hierarchical parameters
\item Gradient-based nested optimization algorithm
\item Errors estimated via curvature as in MLE
\vfill
\item Design complete; awaiting parameter tagging
\end{itemize}

\sld{``Black Box'' VB and EP}
\begin{itemize}
\item \myemph{Black box} so can run any model
\\
{\footnotesize (Laplace or other approximations)}
\item Stochastic, data-streaming \myemph{variational Bayes} (VB)
\item Data-parallel \myemph{expectation propagation} (EP)
\\ {\footnotesize (cavity distributions provide general shard combination)}
\item Both use point estimates of parametric approximation to posterior
\item Optimize parameters to minimize KL divergence
\\
{\footnotesize (VB, EP measure divergence in opposite directions)}
\vfill
\item Prototype stage
\end{itemize}

>>>>

\begin{document}
\sf%
\vspace*{6pt}
%
\noindent
\spc{\huge\bfseries \color{MidnightBlue}{Stan}}
\\[8pt]
\spc{\Large\bfseries \color{MidnightBlue}{Probabilistic Programming Language}}
\\
\vfill
\noindent
\spc\spc{\slshape\footnotesize Core Development Team:}
\\[2pt]
\spc\spc{\small Andrew Gelman, \  \myemph{Bob Carpenter}, \  Matt Hoffman}
\\
\spc\spc{\small Daniel Lee, \  Ben Goodrich, \  Michael Betancourt, \ }
\\
\spc\spc{\small Marcus Brubaker, \   Jiqiang Guo, \  Peter Li, \ }
\\
\spc\spc{\small Allen Riddell, \  Marco Inacio, \ Jeff Arnold, \ }
\\
\spc\spc{\small Mitzi Morris \ }
\vfill
\vfill
\mbox{ } 
\hfill
\includegraphics[width=27.5pt]{img/stanlogo-main.pdf}\hspace*{9pt}
\\
\spc{\footnotesize Microsoft Research NYC 2014}
\hfill 
{\footnotesize \tt mc-stan.org}

\mypart{{\LARGE The Present:} \\ \spc\spc{\Huge Stan 2.4}}

@@

\begin{itemize}
\item Stan is an \myemph{imperative} probabilistic programming language
\vspace*{-12pt}
\begin{itemize}\footnotesize
\item  cf., BUGS: declarative; \ Church: functional; \ Figaro: object-oriented
\end{itemize}
\item Stan \myemph{program}
\vspace*{-4pt}
\begin{itemize}\small
\item declares variables
\item codes log posterior (penalized likelihood)
\item calculates gradients, Hessians, etc., using auto-diff
\end{itemize}
\item Stan \myemph{inference}
\vspace*{-4pt}
\begin{itemize}\small
\item MCMC for full Bayesian inference (+ posterior analysis)
\item MLE for point estimates (+ Hessian-based error estimates)
\end{itemize}
\end{itemize}

\sld{Stan is Open}

\begin{itemize}
\item Stan is \myemph{open source}
\vspace*{-4pt}
\begin{itemize} 
\small
\item Stan Core C++, CmdStan interface: {\slshape new BSD}
\item PyStan, RStan interfaces: {\slshape GPLv3}
\item Dependencies: Eigen, Boost, (googletest)
\end{itemize}
\item GitHub \myemph{hosted publicly} {\small ({\tt stan-dev})}
\vspace*{-4pt}
\begin{itemize}
\small
\item issue tracking (bug reports, feature requests)
\item pull requests, code review, continuous integration hooks
\end{itemize}
\item Google Groups \myemph{public mailing lists} {\small ({\tt stan-dev}, {\tt stan-users})}
\end{itemize}

@@

\sld{Who's Using Stan?}
\begin{itemize}
\item 830+ registrations for mailing list
\item spanning physical, biomedical, and social sciences, 
\\ plus engineering, education, finance, and marketing

\item Papers citing Stan on {\slshape Google Scholar} in subjects
\\[3pt]
{\footnotesize clinical drug
trials, general computational statistics, entomology, opthalmology,
neurology, sociology and population dynamics, genomics, agriculture,
psycholinguistics, molecular biology, population dynamics, materials
engineering, botany, astrophysics, oceanography, election prediction,
fisheries, cancer biology, public health and epidemiology, population
ecology, collaborative filtering for recommender systems, climatology,
educational testing, natural language processing, econometrics}
\end{itemize}

\sld{Documentation and Examples}
%
\begin{itemize}
\item 400+ page core language tutorial manual
\item separate installation, getting started, interface manuals 
\\ {\footnotesize  (RStan, PyStan, CmdStan)}
\item BUGS and JAGS examples (all 3 volumes), 
\item Gelman and Hill, {\slshape Data Analysis Using Regression and
    Multilevel/Hierarchical Models}
\item Wagenmakers and Lee, {\slshape Bayesian Cognitive Modeling}
\item two other books in progress
\item user-contributed examples in group
\end{itemize}

\sld{Scaling and Evaluation}
\begin{itemize}
\item Type of scaling \hfill (improvement$^{*}$)
\vspace*{-4pt}
\begin{itemize}\small
\item more data \hfill (1--2 orders of magnitude)
\item more parameters \hfill (1--2 orders of magnitude)
\item \myemph{more complex models} \hfill ({\large \myemph{$\infty$}})
\end{itemize}
\item imporvement vs.\ ``state-of-the-art'' \myemph{black-box MCMC}
measured by
\vspace*{-4pt}
\begin{itemize}\small
\item time to convergence
\item time per effective sample at mixing
\item memory usage
\end{itemize}
\end{itemize}

@@

\begin{itemize}
\item \myemph{\tt\bfseries data} \ (once) 
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: declare data types, sizes, and constraints
\item {\slshape execute}: read from data source, validate constraints
\end{itemize}
%
\item \myemph{\tt\bfseries parameters} \ (every log prob eval)
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: declare parameter types, sizes, and constraints
\item {\slshape execute}: transform to constrained, optional Jacobian
\end{itemize}
%
\item \myemph{\tt\bfseries model} \ (every log prob eval) 
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: statements definining posterior density
\item {\slshape execute}: execute statements
\end{itemize}
\end{itemize}

@@

\sld{User-Defined Functions}

@@

\begin{itemize}
\item Posterior \myemph{mode finding} via L-BFGS optimization
\\ {\footnotesize (uses model gradient, efficiently approximates Hessian)}
\item \myemph{Disables Jacobians} for parameter inverse transforms
\item Models, data, initialization as in MCMC
\item Curvature (Hessian) used to estimate posterior covariance (or
  standard errors on unconstrained scale)
\end{itemize}

@@

\begin{itemize}
\item No-U-turn sampler (\myemph{NUTS})
\item Subtle algorithm to maintain \myemph{detailed balance}
\item Move randomly \myemph{forward or backward in time}
\item Double number of leapfrog steps each move (\myemph{binary tree})
\item Stop when a subtree makes a \myemph{U-turn}
\\ {\footnotesize (rare: throw away second half if not end to end U-turn)}
\item \myemph{Slice sample} points along last branch of tree
\item \myemph{Generalized} to Riemannian-manifold HMC
\\ {\footnotesize (Betancourt arXiv paper)}
\end{itemize}

@@

\mypart{{\LARGE The Future:} \\ \spc\spc{\Huge Stan 3 \& Beyond}}

\sld{Stan 3: The Refactoring}
\begin{itemize}
\item Third time's the charm in software (re)design
\item Goal of removing duplicated code in interfaces
\item Minimizing dependencies between interfaces and core
\item More modular and flexible
\vspace*{-4pt}
\begin{itemize}\small
\item user-facing C++ API
\item interface API
\end{itemize}
\vfill
\item Design and some coding underway
\end{itemize}

\sld{Differential Equation Solver}
\begin{itemize}
\item Auto-diff solutions w.r.t.\ parameters
\item Integrate coupled system for solution with partials
\item Auto-diff coupled Jacobian for stiff systems
\vfill
\item C++ prototype integrated for large PK/PD models
\vspace*{-4pt}
\begin{itemize}\footnotesize
\item 
Project with Novartis: 
  longitudinal clinical trial w.\ multiple drugs, dosings, placebo control,
  hierarchical model of patient-level effects, meta-analysis
\end{itemize}
\vfill
\item Generalized code complete, under testing
\item {\footnotesize With: Frederic Bois, Amy Racine-Poon, Sebastian Weber}
\end{itemize}

\sld{MLE/MML Error Estimation}
\begin{itemize}
\item Use curvature to estimate error on unconstrained scale
\item Transform back to constrained scale 
\item May require sampling from Laplace approximation
\vfill
\item Design complete
\end{itemize}

@@

\sld{Riemannian Manifold HMC}
\begin{itemize}
\item Supports posteriors with \myemph{position-dependent
    curvature} with local \myemph{mass matrix estimation}
\\
  {\small (e.g., hierarchical models)}
\item \myemph{NUTS} generalized to RHMC
\item \myemph{SoftAbs} metric
\vspace*{-4pt}
\begin{itemize}\footnotesize
\item Eigendecompose Hessian
\item \myemph{positive definite} with positive eigenvalues
\item \myemph{condition} by narrowing eigenvalue range
\end{itemize}
\vfill
\item Code complete; awaiting full higher-order auto-diff
\item {\small (Betancourt {\slshape arXiv} papers)}
\end{itemize}

\sld{Thermodynamic Sampler}
\begin{itemize}
\item Enhances posterior mode-finding with \myemph{multiple modes}
\item Physically motivated alternative to ``simulated''
  \myemph{annealing and tempering} (not really simulated!)
\item Supplies external \myemph{heat bath}
\item Operates through \myemph{contact manifold}
\item System relaxes more naturally between energy levels
\vfill
\item Prototype complete
\item {\small (Betancourt {\slshape arXiv} paper)}
\end{itemize}

\sld{\Large Marginal Maximum Likelihood (MML)}
\begin{itemize}
\item Enable MLE-like estimate with \myemph{infinite likelihoods}
\\ {\small (e.g., hierarchical models, mixture models / clustering)}
\item Marginalize out lower-level parameters to estimate hierarchical
\item Estimate lower-level parameters based on hierarchical estimates
\item Gradient-based nested optimization algorithm
\item Errors / posterior variance estimated as in MLE
\vfill
\item Design complete; awaiting parameter tagging
\end{itemize}

\sld{Variational Bayes (VB)}
\begin{itemize}
\item \myemph{Black box} to \myemph{scale} arbitrary models
\item Batch or stochastic (data streaming)
\item Compute expectations via approximations (e.g., Laplace)
\item Point estimate parametric approximations to posterior
\item Optimize parameters to minimize KL divergence
\vfill
\item Prototype stage 
\item {\footnotesize With: Dave Blei, Alp Kucukelbir, and Rajesh Ranganath}
\end{itemize}

\sld{Expectation Propagation (EP)}
\begin{itemize}
\item \myemph{Black box} to \myemph{scale} arbitrary models
\item Data-parallelization
\vspace*{-4pt}
\begin{itemize}\small
\item cavity distributions guide shard combination
\end{itemize}
\item Point estimate parametric approximations to posterior
\item Optimize parameters to minimize KL divergence
\vfill
\item Prototype stage
\item {\footnotesize With: Aki Vehtari, Nicolas Chopin, Christian Robert,
  John Cunningham}
\end{itemize}
